Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, showing the direction of causal
influence. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. Moreover, they often omit exogenous variables
for simplicity, following a \emph{standard representation}. However,
this section adopts the \emph{magnified representation} of DAGs, as
including exogenous variables in their representation can enhance
clarity and reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the causal relationships among
variables based on Thurstone's theory and the practical aspects of CJ
experiments. Particularly, the graph shows how the texts' discriminal
processes \((T_{ha}, T_{ib})\) influence their discriminal difference
\((D_{hiabjk})\), which in turn determines the outcome. It also depicts
the judges' biases \((B_{kj})\) as an additional factor influencing the
texts' discriminal difference. Furthermore, the DAG helps distinguish
between observed endogenous variables, such as the outcome, and latent
endogenous variables, which include the texts' discriminal processes,
their discriminal difference, and the judges' biases.

@@

\[
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} judge-related variables \((Z_{J})\) causally
influence the judges' bias \((B_{J})\). The DAG also accounts for
idiosyncratic errors \((e_{J})\), capturing the variations in the
judges' bias that relevant judge-related variables cannot explain. Here,
\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
observed matrix with \(n_{J}\) rows and \(q_{J}\) columns. Notably, all
variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} constitute the set of endogenous variables \(V\)
and structural equations \(F\) for the conceptual-population model.

Overall, the conceptual-population model extends Thurstone's theory by
introducing key innovations to address the limitations discussed in
Section~\ref{sec-theory-issues}. These enhancements include accounting
for judges' biases and integrating hierarchical structural components.
Nevertheless, despite its statistical and measurement advantages, the
model still relies on the unrealistic assumption that researchers can
access data capturing all repeated judgments from every judge for each
pair of stimuli produced by every individual in the population. Since
this assumption is rarely met in practice, a more realistic scenario
must be considered.

@@

Considering the sampling mechanism

@@

\[
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

\caption{\label{fig-cj11}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

\caption{\label{fig-cj13}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

>>>>

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, indicating the direction of causal
influence, while the absence of an edge implies no relationship between
the nodes. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally represent observed variables
with solid black circles and unobserved (latent) variables with open
circles \citep{Morgan_et_al_2014}. They also display conditioned
variables by surrounding them with a square. Here, \emph{conditioning}
refers to restricting the statistical analysis to a subset of the
population defined by the values of the given variable
\citep[pp.~32]{Neal_2020}. While DAGs often omit exogenous variables for
simplicity, following a \emph{standard representation}, this section
adopts the \emph{magnified representation} of DAGs, as including
exogenous variables can improve clarity and help reveal potential issues
related to conditioning and confounding \citep{Cinelli_et_al_2020}.

Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the causal relationships among
variables based on Thurstone's theory and the practical aspects of CJ
experiments. Particularly, the graph shows how the texts' discriminal
processes \((T_{ha}, T_{ib})\) influence their discriminal difference
\((D_{hiabjk})\), which in turn determines the outcome. It also depicts
the judges' biases \((B_{kj})\) as an additional factor influencing the
texts' discriminal difference. Furthermore, the DAG distinguish between
observed endogenous variables, such as the outcome, and latent
endogenous variables, which include the texts' discriminal processes,
their discriminal difference, and the judges' biases.

@@

\[
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} \\
\end{aligned}
\]

@@

Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} judge-related variables \((Z_{J})\) causally
influence the judges' bias \((B_{J})\). The DAG also accounts for
idiosyncratic errors \((e_{J})\), capturing the variations in the
judges' bias that relevant judge-related variables cannot explain. Here,
\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
observed matrix with \(n_{J}\) rows and \(q_{J}\) columns.

Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} define the set of endogenous variables \(V\),
structural equations \(F\), and exogenous variables \(E\) for the
conceptual-population model. Additionally, the figures demonstrate that
all exogenous variables are independent of one another, as indicated by
the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\).

Overall, the conceptual-population model extends Thurstone's theory by
introducing key innovations to address the limitations discussed in
Section~\ref{sec-theory-issues}. These enhancements include accounting
for judges' biases and integrating hierarchical structural components.
Nevertheless, despite its promise of enhancing measurement accuracy and
precision, the model still depends on the unrealistic assumption that
researchers have access to data encompassing all repeated judgments from
every judge for each pair of stimuli produced by every individual in the
population. Since researchers rarely meet this assumption in practice,
they must consider a more realistic scenario. The extensions to the
conceptual-population model, derived from this realistic scenario, are
concisely described in the sample-comparison model.

@@

In the sample-comparison model, researchers have access to a limited
number of judgments (\(n^{s}{K}\)) from a sample of judges
(\(n^{s}{J}\)), along with a specific number of texts (\(n^{s}{A}\))
from a sample of students (\(n^{s}{I}\)) drawn from the conceptual
population.

Figure~\ref{fig-cj11} illustrates the sample-comparison model,
incorporating the sample mechanism. Following the convention of
\citet[pp.~499-516]{McElreath_2020}, the DAG~\ref{fig-cj11_dag} shows
several conditioned variables \((S_{R}\), \(S_{P}\), \(S_{U}\), and
\(S_{V})\) that causally influence the sample-comparison outcome
\(O^{sc}_{RPUV}\), reflecting how the outcome is a sample drawn from the
conceptual population. The DAG also depicts the conceptual-population
outcome \(O^{cp}_{RPUV}\) is now an unobserved variable, emphasizing the
researchers' inability to access such an outcome.

@@

\[
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} \\
\end{aligned}
\]

@@

\caption{\label{fig-cj11}Sample-comparison model including sampling
mechanism}

@@

\[
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} \\
\end{aligned}
\]

@@

\caption{\label{fig-cj13}Sample-comparison model including sample and
comparison mechanisms}
